---
title: "p8105_hw2_zj2379"
author: "Zheshu Jiang"
date: "2023-10-02"
output: github_document
---

##Problem 1
First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.
```{r}
library(tidyverse)
```

```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )
```

Clean the data in pols-month.csv
```{r}
pols=
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv",show_col_types = FALSE) %>%
  janitor::clean_names() %>% 
  separate(mon,into=c("year","month","day"),sep="-") %>% 
  mutate_at(vars(year, month, day), as.numeric) %>% 
  mutate(month=recode(month, '01'="January" , '02'="February",'03'="March",'04'="April","05"="May","06"="June","07"="July","08"="August","09"="Septemer","10"="October","11"="November","12"="December")) %>% 
   mutate(president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) %>% 
  select(year, month, everything(), -day, -starts_with("prez"))
```
Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") %>% 
  janitor::clean_names() %>% 
  separate(date, into = c("month", "day", "year"), convert = TRUE) %>% 
  arrange(year, month) %>% 
  mutate(month = month.name[month]) %>% 
  select(year, month, close) 
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") %>%
   rename(year = Year) %>% 
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) %>% 
  select(year, month, unemployment)
```
Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r merge_538}
data_538 = 
  left_join(pols, snp)|>
  left_join(x = _, y = unemployment)
```

## problem 2

```{r}
library(readxl)
```

```{r}
mrtrashweel=
  #read the Mr. Trash Wheel sheet
   read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet="Mr. Trash Wheel") %>%
  #clean the dataset and filter out missing values
  janitor::clean_names() %>%filter(if_any(everything(), ~ !is.na(.)))%>%
  #update the homes_powered values
  mutate(homes_powered=(500*weight_tons)/30) %>%  
  #remove desimal points from homes_powered values
  mutate(homes_powered=trunc(homes_powered)) %>% 
  #remove some columns
  select(-x15,-x16) %>% 
  #convert year from character to numeric variable
  mutate_at(vars(year), as.numeric)


mrtrashweel
```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.
```{r}
gwynndatrashweel=
  #read the Gwynnda Trash Wheel sheet
   read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet="Gwynnda Trash Wheel") %>% 
  #clean the dataset and filter out missing values
  janitor::clean_names() %>%filter(if_any(everything(), ~ !is.na(.))) %>% 
  #update the homes_powered values
  mutate(homes_powered=(500*weight_tons)/30) %>%  
  #remove desimal points from homes_powered values
  mutate(homes_powered=trunc(homes_powered))

gwynndatrashweel
```
```{r}
professortrashweel=
  #read the Professor Trash Wheel sheet
   read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet="Professor Trash Wheel") %>% 
  #clean the dataset and filter out missing values
  janitor::clean_names() %>%filter(if_any(everything(), ~ !is.na(.))) %>% 
  #update the homes_powered values
  mutate(homes_powered=(500*weight_tons)/30) %>%  
  #remove desimal points from homes_powered values
  mutate(homes_powered=trunc(homes_powered))

professortrashweel
```


```{r}
#merge three datasets
data_combine = 
  left_join(mrtrashweel,professortrashweel)|>
  left_join(x = _, y = gwynndatrashweel)
data_combine
```


Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?

```{r}
#create a sub-dataset from gwynndajuly only have the observations from july
gwynndajuly <- filter(gwynndatrashweel, month == "July")
#calculate the sum of igarette_butts
sum(gwynndajuly$cigarette_butts) 
```

The `mrtrashweel` data has `r nrow(mrtrashweel)` obervations and `r ncol(mrtrashweel)` variables and tells us about the weight of different types of trash for a given year from years 2014 to 2022 to power the number of homes ranging from `r range(mrtrashweel$homes_powered)[1]` to `r range(mrtrashweel$homes_powered)[2]`. 

The `professortrashweel` data has `r nrow(professortrashweel)` obervations and `r ncol(professortrashweel)` variables and tells us about the weight of different types of trash for a given year from years 2017 to 2022 to power the number of homes ranging from `r range(professortrashweel$homes_powered)[1]` to `r range(professortrashweel$homes_powered)[2]`. 

The `gwynndatrashweel` data has `r nrow(gwynndatrashweel)` obervations and `r ncol(gwynndatrashweel)` variables and tells us about the weight of different types of trash for a given year from years 2021 to 2022 to power the number of homes ranging from `r range(gwynndatrashweel$homes_powered)[1]` to `r range(gwynndatrashweel$homes_powered)[2]`. 

The `data_combine` data has `r nrow(data_combine)` obervations and `r ncol(data_combine)` variables and tells us about the weight of different types of trash for a given year from years 2014 to 2022 to power the number of homes ranging from `r range(data_combine$homes_powered)[1]` to `r range(data_combine$homes_powered)[2]`. 

The total weight of trash collected by Professor Trash Wheel is `r sum(professortrashweel$weight_tons)`tons and the total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(gwynndajuly$cigarette_butts)`.



