---
title: "p8105_hw2_zj2379"
author: "Zheshu Jiang"
date: "2023-10-02"
output: github_document
---

##Problem 1
First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.
```{r}
library(tidyverse)
```

```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )
```

Clean the data in pols-month.csv
```{r}
pols=
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv",show_col_types = FALSE) %>%
  janitor::clean_names() %>% 
  separate(mon,into=c("year","month","day"),sep="-") %>% 
  mutate_at(vars(year, month, day), as.numeric) %>% 
  mutate(month=recode(month, '01'="January" , '02'="February",'03'="March",'04'="April","05"="May","06"="June","07"="July","08"="August","09"="Septemer","10"="October","11"="November","12"="December")) %>% 
   mutate(president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) %>% 
  select(year, month, everything(), -day, -starts_with("prez"))
```
Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") %>% 
  janitor::clean_names() %>% 
  separate(date, into = c("month", "day", "year"), convert = TRUE) %>% 
  arrange(year, month) %>% 
  mutate(month = month.name[month]) %>% 
  select(year, month, close) 
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") %>%
   rename(year = Year) %>% 
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) %>% 
  select(year, month, unemployment)
```
Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r merge_538}
data_538 = 
  left_join(pols, snp)|>
  left_join(x = _, y = unemployment)
```

## problem 2

```{r}
library(readxl)
```

```{r}
mrtrashweel=
   read_excel("./data/202207 Trash Wheel Collection Data.xlsx") %>%
  janitor::clean_names() %>%filter(if_any(everything(), ~ !is.na(.)))%>%
  mutate(homes_powered=(500*weight_tons)/30) %>% 
  mutate(homes_powered=trunc(homes_powered))


mrtrashweel
```

